---
title: "Extending riskmetric"
author: "R Validation Hub"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending riskmetric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(riskmetric)
library(dplyr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(repos = "https://cran.rstudio.com")

print_riskmetric_function <- function(names) {
  for (name in names) {
    fprint <- capture.output(getNamespace("riskmetric")[[name]])
    fprint <- fprint[!startsWith(fprint, "<")]
    cat(name, ' <- ', sep = '')
    cat(paste(fprint, collapse = '\n'), '\n\n')
  }
}
```

`riskmetric` is designed to be readily extensible. This is done through use of
the S3 method dispatch system and a consciou acknowledgement of the varying
needs that someone may have when assessing package risk. With this in mind,
every user facing function is designed first and foremost to be flexible.

Here we'll walk through a trivial example where we'll extend `riskmetric` to add
a new assessment, scoring and risk summary function to determine the risk
associated with a package given its name starts with the letter "r".

# Adding an Assessment

Assessments are the atomic unit of the `riskmetric` package, and are used to
kick off an individual metric evaluation. Each assessment is a generic function
starting with an `assess_` prefix, which can dispatch based on the subclass of
the `pkg_ref` object.

## Assessment Example

As an example, take a look at how `assess_has_news` has been implemented. We'll
focus on just the generic and the `pkg_install` functions:

```{r, echo = FALSE}
print_riskmetric_function(c("assess_has_news", "assess_has_news.pkg_install"))
```

There are a couple things to note. First, the S3 system is used to dispatch
functionality for the appropriate package reference class. Since the way we'd
assess the inclusion of a NEWS file might be different for an installed package
or remotely sourced metadata, we may have distinct functions to process these
datatypes in distinct ways.

Second, a cosmetic `"column_name"` attribute is used, which replaces the
function name as the name of the new column after using the `assess()` verb.

Finally, a `pkg_metric` object is returned, which stores arbitrary data
pertaining to the metric and importantly adopts a unique subclass for the
assessment function.

## Writing a New Assessment

Now we'll write our assessment. Eventually we want to consider a package high
risk if the name does not start with "r". We'll need to make a `pkg_metric`
object containing the first letter of the name.

```{r, results = 'hide'}
assess_name_first_letter <- function(x, ...) {
  UseMethod("assess_name_first_letter")
}
attr(assess_name_first_letter, "column_name") <- "name_first_letter"

assess_name_first_letter.pkg_ref <- function(x, ...) {
  new_pkg_metric(substr(x$name, 0, 1), class = "pkg_metric_name_first_letter")
} 
```

# Adding `pkg_ref` Metadata

Perhaps we want to reuse metadata used when assessing the first letter so that
it can be reused by other assessments. For particularly taxing metadata, such as
metadat that requires a query against a public API, scraping a web page or a
large data download, it's important to store it for other assessment functions
to reuse.

To handle this, we define a function for `pkg_ref_cache` to dispatch to.

## Example Metadata Caching

This is how the `riskmetric` package handles parsing the DESCRIPTION file so
that it can feed into all downstream assessments without having to re-parse the
file each time or copy the code to do so.

```{r, echo = FALSE}
print_riskmetric_function(c("pkg_ref_cache.description", "pkg_ref_cache.description.pkg_install"))
```

Once these are defined, they'll be automatically called when the field is first
accessed by the `pkg_ref` object, and then stored for any downstream uses.

```{r, eval = FALSE}
library(riskmetric)
package <- pkg_ref("riskmetric")
```

```{r, echo = FALSE}
package_real <- pkg_ref("../../riskmetric")
```

```{r, echo = FALSE}
rver <- gsub("\\.\\d+$", "", paste0(R.version$major, ".", R.version$minor))
package <- pkg_ref("riskmetric")

# hack in order to mutate package environment directly so nobody accidentally
# publishes any personal info in their library path
invisible(riskmetric:::bare_env(package, {
  package$path <- sprintf("/home/user/username/R/%s/Resources/library/riskmetric", rver)
}))

package
```

Notice that upon initialization, the `description` field indicates that it
hasn't yet been evaluated with a trailing `...` in the name.

```{r, eval = FALSE}
str(package$description)
```

```{r, echo = FALSE}
str(package_real$description)
```

Now, when we look at the contents of our `pkg_ref` object, we see that the
`DESCRIPTION` field is completed.

```{r, eval = FALSE}
package
```

```{r, echo = FALSE}
package_real
```

Because the `pkg_ref` object stores an environment, caching this values once
makes them available for any future attempts to access the field. This is
helpful because we, as developers of the package, don't need to think critically
about the order that assessments are performed, and allows users to redefine the
order of assessments without worry about how metadata is acquired.

## Writing a Metadata Cache

Now, for our new metric, we want to cache the package name's first letter. We
need to add a new `pkg_ref_cache` function for our field. Thankfully, any
subclass of `pkg_ref` can access the first letter the same way, so we just need
the one function.

```{r}
pkg_ref_cache.name_first_letter <- function(x, name, ...) {
  substr(x$name, 0, 1)
}
```

After adding this caching function, we need to make a small modification to
`assess_name_first_letter.pkg_ref` in order use our newly cached value.

```{r}
assess_name_first_letter.pkg_ref <- function(x, ...) {
  new_pkg_metric(x$name_first_letter, class = "pkg_metric_name_first_letter")
} 
```

Let's try it out!

```{r}
package$name
package$name_first_letter
```

# Defining an Assessment Scoring Function

Next, we need a function for scoring our assessment output. In this case, our
output is a `pkg_metric` object whose data is the first letter of the package
name.

We'll add a dispatched function for the `score` function. As a convention, these
functions return a numeric value representing how well the package conforms to
best practices with values between 0 (poor practice) and 1 (best practice).

```{r}
score.pkg_metric_name_first_letter <- function(x, ...) {
  as.numeric(x == "r")
}
```

# Adding our Assessment to the `assess()` Verb

The `assess` function accepts a list of functions to apply. `riskmetric`
provides a shorthande, `all_assessments()` to collect all the included
assessment functions, and you're free to add to that list to customize your own
assessment toolkit.

```{r}
library(dplyr)
pkg_ref(c("riskmetric", "utils", "tools")) %>%
  as_tibble() %>%
  assess(c(all_assessments(), assess_name_first_letter))
```

Our scoring function will automatically get picked up and used by the score
method.

```{r}
pkg_ref(c("riskmetric", "utils", "tools")) %>%
  as_tibble() %>%
  assess(c(all_assessments(), assess_name_first_letter)) %>%
  score()
```

and we can define our own summarizing function to aggregate our scores into a
single numeric risk.

```{r}
pkg_ref(c("riskmetric", "utils", "tools")) %>%
  as_tibble() %>%
  assess(c(all_assessments(), assess_name_first_letter)) %>%
  score() %>%
  mutate(risk = (has_news + name_first_letter) / 2)
```

# How you can help...

The `riskmetric` package was designed to be easily extensible. You can develop
dispatched functions in your development environment, hone them into well formed
assessments and contribute them back to the core `riskmetrics` package once
you're done.

If you'd like feedback before embarking on developing a new metric, please feel
free to [file an issue on the riskmetric GitHub](https://github.com/pharmaR/riskmetric/issues/new?labels=Metric%20Proposal). 
